---
title: "Introduction to mixexptr"
output: rmarkdown::html_vignette
description: |
  Analyze microbial interactions in mix experiments
vignette: >
  %\VignetteIndexEntry{Introduction to mixexptr}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

`mixexptr` is an R package for analyzing interactions among microbes. 
A common experimental design is to mix together two different microbes 
(different bacterial genotypes, for example) then measure how their fitness and 
behavior depends on mix frequency. How do they behave differently together 
compared to on their own? 
`mixexptr` provides tools to calculate and plot the fitness effects in these mix 
experiments so that researchers can get the most out of their data. 

## Calculate fitness effects

```{r setup}
library(mixexptr)
```

`mixexptr` provides a convenient way to calculate best-practice fitness measures that are:

* Robust and quantitatively comparable across different species and types of interaction
* Well-suited to statistical analysis of effect sizes and confidence intervals
* Meaningful for both kin and multilevel selection theories of social evolution

```{r}
fitness_results <- calculate_mix_fitness(
	data_smith_2010, 
	var_names = c(
		initial_number_A = "initial_cells_evolved",
		initial_number_B = "initial_cells_ancestral",
		final_number_A = "final_spores_evolved",
		final_number_B = "final_spores_ancestral", 
		name_A = "GVB206.3", 
		name_B = "GJV10"
	)
)
head(fitness_results)
```

## Plot fitness effects

`mixexptr` provides convenient ways to plot calculated fitness effects. Here's a quick diagnostic plot of the different fitness measures for this dataset: 
```r
plot_mix_fitness(fitness_results, mix_scale = "fraction")
```

## Further reading

* smith j and Inglis RF (2021) Evaluating kin and group selection as tools for quantitative analysis of microbial data. Proceedings B 288:20201657. [https://doi.org/10.1098/rspb.2020.1657](https://doi.org/10.1098/rspb.2020.1657)

