---
title: "Analyzing microbial interactions with mixexptr"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Analyzing microbial interactions with mixexptr}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

A common experimental design in microbiology is to mix together two different 
microbes (different genotypes of bacteria, for example) then measure how their 
behavior depends on mix frequency. 
How do they act differently together than when they're on their own? 
How does mixing affect their survival and reproductive success (their fitness)?

mixexptr is an R package for analyzing these mix experiments. 
It helps researchers get the most out of their data by providing tools to 
calculate and plot fitness measures that are: 

* Robust and quantitatively comparable across different species and types of 
  interaction
* Meaningful for both kin and multilevel selection theories of social evolution
* Well-suited to statistical analysis of effect sizes and confidence intervals


<!-- TEACH PEOPLE HOW TO USE THE PACKAGE -->

The basic workflow for analyzing mix experiments with mixexptr is: 

1. Make a data frame containing the experimental observations used to calculate 
   fitness
2. Calculate fitness measures with `calculate_mix_fitness()`
3. Inspect a diagnostic plot of fitness measures with `plot_mix_fitness()`
4. Plot the measures of interest with `plot_strain_fitness()`, 
  `plot_total_group_fitness()`, and/or `plot_within_group_fitness()`
5. Fit statistical models of your choosing and plot them with the data 


## Input data

The input format for mixexptr is a data frame where each row describes microbial abundance at the beginning and end of a mix experiment. The experimentally observed quantities could be any combination of strain abundance (number or density), strain frequency (fraction of total), or total group abundance---anything sufficient to define the absolute abundance of both strains in the experiment. Each row should contain data for a single observational unit: a single experimental replicate with one strain combination, one mix frequency, one combination of experimental treatments, etc.  

For example, let's look at some datasets included with mixexptr. This dataset from smith *et al.* (2010) is an experiment mixing a wild-type strain of *Myxococcus* bacteria with an experimentally-evolved cheater strain. The observed quantities are the number of cells from each strain before and after they form multicellular fruiting bodies. 

```{r data-smith-2010}
library(mixexptr)
head(data_smith_2010)
```

This dataset from Yurtsev *et al.* (2013) describes the results of mix experiments with an antibiotic-sensitive strain of *Escherichia coli* and an antibiotic-resistant strain that detoxifies its local environment. The observed quantities here are total cell density (from OD~600~) and the fraction of cells belonging to each strain (from flow cytometry). The dataset also includes several experimental treatments with different antibiotic concentrations and different amounts of culture growth. 

```{r data-Yurtsev-2013}
head(data_Yurtsev_2013)
```


## Calculating fitness

`calculate_mix_fitness()` is the function to calculate fitness. It accepts an input data frame of microbial abundances and a character vector describing what the data columns are. It returns a data frame with several calculated measures of fitness and two measures of mix frequency. For example: 

```{r fitness_smith_2010}
fitness_smith_2010 <- calculate_mix_fitness(
	data = data_smith_2010, 
	var_names = c(
		initial_number_A = "initial_cells_evolved",
		initial_number_B = "initial_cells_ancestral",
		final_number_A = "final_spores_evolved",
		final_number_B = "final_spores_ancestral", 
		name_A = "GVB206.3", 
		name_B = "GJV10"
	)
)
head(fitness_smith_2010)
```

Use the `keep` argument to include data columns for experimental treatment, experimental block, and so on in the output frame. For the Yurtsev data, for example, we want to keep antibiotic concentration and culture dilution: 

```{r fitness_Yurtsev_2013}
fitness_Yurtsev_2013 <- calculate_mix_fitness(
	data_Yurtsev_2013, 
	var_names = c(
		initial_number_total = "OD_initial",
		initial_fraction_A = "fraction_resistant_initial",
		final_number_total = "OD_final",
		final_fraction_A = "fraction_resistant_final",
		name_A = "AmpR",
		name_B = "AmpS"
	), 
	keep = c("ampicillin", "dilution")
)
head(fitness_Yurtsev_2013)
```

Note the warning messages here. `calculate_mix_fitness()` will warn you if any of the data are not biologically meaningful. There cannot be a negative number of cells, for example, and fractions are bounded at zero and one. Some of the Yurtsev data includes negative fractions---likely an artifact of correcting for background fluorescence in flow cytometry. 

* Explain `name_A` and `name_B` arguments
* Make explicit which is which in the output itself
* Column or label
* Column example: Madgwick data



## Fitness measures

The fitness measures calculated by `calculate_mix_fitness()` are:

* `fitness_A` : Fitness of strain A
* `fitness_B` : Fitness of strain B
* `fitness_total` : Total-group fitness (strains A and B combined)
* `fitness_ratio_A_B` : Within-group ratio of strain A fitness to strain B fitness

These fitness measures have been chosen because ...

* Robust across systems and types of interaction
* Strong, nonadditive effects
* Both strain and group-focused fitness outcomes
* Wrightian fitness ... easy intepretation as fold change in abundance ... for example, ...

The fitness calculations are very simple. If the initial abundance of strain A (measured as the number or density of cells or virions) is $n_A$ and its final abundance is $n'_A$, then its fitness is $w_A = n'_A/n_A$. Similarly, the fitness of strain B is $w_B = n'_B/n_B$. Total group fitness is $W = (n'_A + n'_B)/(n_A + n_B)$ and the within-group fitness ratio is $w_A/w_B$. 

The two measures of mix frequency calculated by `calculate_mix_fitness()` are:

* `initial_fraction_A` : $n_A / (n_A + n_B)$  
* `initial_ratio_A_B` : $n_A / n_B$ 


## Plotting fitness

mixexptr offers several functions for plotting fitness measures: 

* `plot_mix_fitness()` diagnostic plot
* `plot_strain_fitness()`
* `plot_total_group_fitness()`
* `plot_within_group_fitness()`

**Explain plot functions**

* Wrightian fitness best visualized and analyzed over log scales
* Issue: zeroes
* x-axis quantitative ... fitness effects of microbial interactions usually depend on mix frequency
* `initial_fraction_A` ... linear scale from zero to one
* `initial_ratio_A_B` ... log scales ... better for visualizing effects at ... can't show single-strain experiments

**Extending plots**

These are convenience functions that make plots with the ggplot2 package using with reasonable default settings. They return ggplot2 objects that you are free to extend and modify, adding or replacing scales, themes, and so on. 

* Including fitted statistical models
* Combine subplots with ??? package


## Other stuff (maybe include above)

* mixexptr functions work with magrittr pipe `%>%` and base R pipe `|>`


## References

* smith j, Van Dyken JD, and Zee PC (2010) A generalization of Hamilton's rule for the evolution of microbial cooperation. *Science* **328:** 1700-1703. 
[https://doi.org/10.1126/science.1189675](https://doi.org/10.1126/science.1189675)

* smith j and Inglis RF (2021) Evaluating kin and group selection as tools for 
quantitative analysis of microbial data. *Proceedings B* **288:** 20201657.
[https://doi.org/10.1098/rspb.2020.1657](https://doi.org/10.1098/rspb.2020.1657)

* Yurtsev EA, Chao HX, Datta MS, Artemova T, and Gore J (2013) Bacterial cheating drives the population dynamics of cooperative antibiotic resistance plasmids. *Molecular Systems Biology* **9:** 683. 
[https://doi.org/10.1038/msb.2013.39](https://doi.org/10.1038/msb.2013.39)
